---
title: "Getting started with bcsv"
format:
  html:
    df-print: kable
vignette: >
  %\VignetteIndexEntry{Getting started with bcsv}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`bcsv` reads and writes CSV files paired with a **bcsv** metadata sidecar—a
JSON-LD extension of W3C CSVW that adds ordered/categorical factors, units,
multiple NA codes, and a SHA-256 integrity hash. Files come in pairs by
convention: `data.csv` ↔ `data.json`.

Install:

```r
install.packages("bcsv")
```

Each recipe below stands on its own. The setup chunk defines `example_path()`,
which returns the path to a dataset shipped inside the package (the cookbook
uses `course_feedback`).

```{r setup}
library(bcsv)
example_path <- function(name) system.file("extdata", name, package = "bcsv")
cat("schema", bcsv::SCHEMA_VERSION, "\n")
```

## Load a typed dataset

`read_bcsv` applies the declared types: `rating` becomes an **ordered** factor,
`date` a `Date`, `passed` logical.

```{r}
csv_path <- example_path("course_feedback.csv")
json_path <- example_path("course_feedback.json")

df <- read_bcsv(csv_path, json_path)
str(df)
```

Compare with `read.csv`, which sees only text and numbers—`rating` and `date`
arrive as untyped character columns:

```{r}
df_plain <- read.csv(csv_path)
str(df_plain)
```

Every column's declared metadata—types, units, levels, descriptions—travels
in the sidecar. `tabularize()` (alias `tabularise()`) renders it as a table
(units live here, not on the typed column):

```{r}
metadata <- jsonlite::fromJSON(json_path, simplifyVector = FALSE)
tabularize(metadata)
```

## Ordered factors: why declared levels matter

The payoff of the `ordered` datatype is clearest when the natural order is *not*
alphabetical—like a satisfaction rating. A plain read sees `rating` as text
and can only sort it alphabetically, which is meaningless:

```{r}
sort(unique(read.csv(csv_path)$rating))
```

Because the metadata declares `rating` as `ordered` with explicit `levels`,
`read_bcsv` rebuilds the real scale—so sorting and comparisons follow it:

```{r}
levels(df$rating)
```

```{r}
df[df$rating >= "good", c("name", "rating")]
```

## Validate a pair

`validate_bcsv` returns `valid`, plus `errors` / `warnings` lists and the
columns it compared—it **collects** issues rather than stopping. The result
prints itself as a summary (a `print` method built on `tabularize()`); call
`tabularize(res)` when you want the issues as a data frame. A clean pair has
no issues:

```{r}
csv_path <- example_path("course_feedback.csv")
json_path <- example_path("course_feedback.json")

res <- validate_bcsv(csv_path, json_path)
res
```

Now a dataset that violates a declared constraint—`score` is capped at 100 but
one row is 150. The violation is collected as a warning (so `valid` stays `TRUE`;
`on_violation = "raise"` would promote it to an error):

```{r}
tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
bad_csv <- file.path(tmpdir, "scores.csv")
readr::write_csv(data.frame(id = 1:2, score = c(88, 150)), bad_csv)

scores_metadata <- document_bcsv(
  data.frame(id = 1L, score = 0.0),
  "Exam scores.",
  data_url = "scores.csv",
  column_descriptions = list(score = list(maximum = 100))
)
jsonlite::write_json(scores_metadata, file.path(tmpdir, "scores.json"),
                     auto_unbox = TRUE, null = "null")

res <- validate_bcsv(bad_csv, check_hash = FALSE)
res$valid
tabularize(res)
```

## Multiple NA codes

A column can declare extra `na_strings`; the empty string is always NA. They all
collapse to `NA` uniformly.

Here is a CSV whose `score` column uses several NA tokens:

```{r}
tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
csv <- file.path(tmpdir, "scores.csv")
readr::write_csv(data.frame(id = 1:4, score = c("88", "NA", "-99", "")), csv)
```

You can write the metadata by hand, or scaffold it from the column types with
`document_bcsv`—which fills in the field names and rejects invalid ones, so
typos are caught. The two produce the same column definitions:

```{r}
by_hand <- list(
  `@context` = bcsv::CONTEXT_URL, url = "scores.csv",
  description = "Exam scores.",
  table_schema = list(columns = list(
    list(name = "id", datatype = "integer"),
    list(name = "score", datatype = "number", na_strings = list("NA", "-99"))
  ))
)
scaffolded <- document_bcsv(
  data.frame(id = 1L, score = 0.0),
  "Exam scores.",
  data_url = "scores.csv",
  column_descriptions = list(score = list(na_strings = list("NA", "-99")))
)
identical(scaffolded$table_schema$columns, by_hand$table_schema$columns)
```

```{r}
jsonlite::write_json(scaffolded, file.path(tmpdir, "scores.json"), auto_unbox = TRUE, null = "null")
df <- read_bcsv(csv, check_hash = FALSE)   # hand-made metadata has no file_hash
df
```

```{r}
is.na(df$score)
class(df$score)
```

## Lenient vs strict reads

`read_bcsv` defaults to `on_violation = "warn"`: a value that fails coercion
becomes `NA` with a warning, and the read returns. `on_violation = "raise"`
stops on the first violation.

```{r}
tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
csv <- file.path(tmpdir, "ages.csv")
readr::write_csv(data.frame(id = 1:3, age = c("25", "thirty", "41")), csv)

meta <- document_bcsv(data.frame(id = 1L, age = 1L), "Participant ages.", data_url = "ages.csv")
jsonlite::write_json(meta, file.path(tmpdir, "ages.json"), auto_unbox = TRUE, null = "null")

df <- withCallingHandlers(
  read_bcsv(csv, check_hash = FALSE),
  warning = function(w) { message("warning: ", conditionMessage(w)); invokeRestart("muffleWarning") }
)
df$age
```

```{r}
tryCatch(
  read_bcsv(csv, check_hash = FALSE, on_violation = "raise"),
  bcsv_validation_error = function(e) paste("strict →", conditionMessage(e))
)
```

## Verify file integrity

The metadata carries a SHA-256 `file_hash`. `read_bcsv` recomputes it and raises
`bcsv_validation_error` on a mismatch—the data has changed since it was
documented. `check_hash` and `check_schema` are independent toggles.

```{r}
csv_path <- example_path("course_feedback.csv")
json_path <- example_path("course_feedback.json")

# The bundled pair is consistent, so it reads without complaint:
feedback <- read_bcsv(csv_path, json_path)
head(feedback, 3)   # consistent pair → reads fine
```

Now edit the data while leaving the metadata's hash stale, and read it back:

```{r}
tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
modified_csv <- file.path(tmpdir, "course_feedback.csv")
file.copy(json_path, file.path(tmpdir, "course_feedback.json"))   # keep the original (stale) hash

data <- read.csv(csv_path)
data$name[1] <- "Alicia"
write.csv(data, modified_csv, row.names = FALSE)

tryCatch(
  read_bcsv(modified_csv),
  bcsv_validation_error = function(e) paste("integrity →", substr(conditionMessage(e), 1, 60))
)
```

```{r}
# When the change is intentional, skip the integrity check to read it anyway:
reread <- read_bcsv(modified_csv, check_hash = FALSE)
head(reread, 3)   # same data, integrity check skipped
```

## Generate metadata for a data frame

`document_bcsv` infers datatypes from the (already typed) columns and layers in
`column_descriptions`. The table-level description is a required
argument—the bcsv schema insists every table describe what it is about.

```{r}
study_data <- data.frame(
  agent_id = 1:3,
  response_time = c(0.45, 0.52, 0.39),
  correct = c(TRUE, FALSE, TRUE)
)
study_metadata <- document_bcsv(
  study_data,
  "Reaction-time pilot data.",
  pretty_name = "Reaction Time Pilot",
  column_descriptions = list(
    response_time = list(unit = "s", minimum = 0, description = "Response time"),
    correct = list(description = "Whether the response was correct")
  ),
  primary_key = "agent_id"
)
```

```{r}
tabularize(study_metadata)
```

Unknown keys are rejected immediately—the allowed key is `unit`, not `units`:

```{r}
tryCatch(
  document_bcsv(study_data, "Reaction-time pilot data.", column_descriptions = list(response_time = list(units = "s"))),
  bcsv_unknown_key_error = function(e) paste("rejected →", substr(conditionMessage(e), 1, 70))
)
```

## Write a pair, read it back

`write_bcsv` writes the CSV and metadata atomically and computes the
`file_hash`. An ordered factor round-trips back ordered.

```{r}
ratings <- data.frame(
  agent_id = 1:3,
  rating = ordered(c("good", "very bad", "very good"), levels = c("very bad", "bad", "good", "very good"))
)
ratings_metadata <- document_bcsv(ratings, "Agent satisfaction ratings.", pretty_name = "Ratings", primary_key = "agent_id")

tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
written <- write_bcsv(ratings, file.path(tmpdir, "ratings.csv"), ratings_metadata)
written
```

```{r}
reloaded <- read_bcsv(written$data)
is.ordered(reloaded$rating)
levels(reloaded$rating)
as.character(reloaded$rating)
```

## Author metadata for a CSV

When you have a bare CSV, the authoring verbs build its metadata through an
editable YAML, keeping a **single metadata document** on disk at a time—YAML
while you edit, JSON once finalized. The directory listing after each step
shows which: if a step fails, nothing is written, so the files on disk always
tell you where you are.

`draft_metadata` writes a YAML prefilled from the data: a commented legend of
the optional column keys, a placeholder for the table description (which the
bcsv schema **requires**), and one placeholder description per column—exactly
what you want when a questionnaire CSV has hundreds of columns named like
`q12r`:

```{r}
tmpdir <- tempfile(); dir.create(tmpdir)   # a throwaway folder for this example's files
raw <- file.path(tmpdir, "survey.csv")
readr::write_csv(data.frame(id = 1:2, score = c(88, 92)), raw)

draft <- draft_metadata(raw)     # writes survey.yaml, prefilled from the data
cat(readLines(draft), sep = "\n")
```

(`draft_metadata` also accepts a data frame: `draft_metadata(df, "survey")`
writes both `survey.csv` and `survey.yaml`.)

Placeholders you never edit are **stripped at finalize time**, so boilerplate
can't leak into published metadata—and because the table description is
required, finalizing before describing the table is refused, writing nothing:

```{r}
res <- suppressMessages(finalize_metadata(raw))   # placeholders untouched → refused
list.files(tmpdir)               # survey.yaml still there, no survey.json
```

This is the point where you would open `survey.yaml` in your editor. Here the
edit is scripted so the vignette runs end-to-end—describe the table, describe
`score` and cap it; `id` keeps its placeholder deliberately:

```{r}
meta <- yaml::read_yaml(draft)
meta$description <- "End-of-course survey scores, spring term."
meta$table_schema$columns <- lapply(meta$table_schema$columns, function(col) {
  if (col$name == "score") { col$description <- "Exam score"; col$maximum <- 100 }
  col
})
yaml::write_yaml(meta, draft)
```

`finalize_metadata` validates the pair, and only on success: standardizes the
CSV into the canonical bcsv dialect (comma delimiter, no row numbers, LF line
endings, UTF-8—byte-identical across R and Python), writes `survey.json`
with a fresh `file_hash`, and removes the YAML:

```{r}
res <- finalize_metadata(raw)
list.files(tmpdir)               # survey.yaml replaced by canonical survey.json
```

The untouched `id` placeholder was stripped rather than published—`id` simply
has no description in the canonical metadata:

```{r}
print_metadata(raw)              # the finalized metadata, as a table
```

Changed your mind? `edit_metadata` converts the JSON back to an editable YAML
(removing the JSON), and another `finalize_metadata` closes the loop:

```{r}
edit_metadata(raw)               # survey.json → survey.yaml, ready to edit
list.files(tmpdir)
```

```{r}
res <- finalize_metadata(raw)    # (further YAML edits would go here)
list.files(tmpdir)
```

`export_metadata` writes a readable copy under a distinct name—for sharing
or review—without disturbing the canonical pair:

```{r}
export <- export_metadata(raw)   # default: YAML; output_format = "json" also works
list.files(tmpdir)
```
